The cost of school consolidations: impacts of reduced accessibility

### 2024-04-26 paper/Manuscript.qmd
manuscript draft for AP review

---
title: "The cost of school consolidations: impacts of reduced accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
linenumbers: true
numbersections: true
format: docx
editor: source
execute: 
  freeze: auto
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: McMaster University SEES
    footnote: "Corresponding Author"
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliation: University of Toronto
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca "
    affiliation: McMaster University CE
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: McMaster University SEES
address:
  - code: McMaster University SEES
    address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - code: McMaster University CE
    address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - code: University of Toronto
    address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"  
abstract: | 
  | The reduction of mobility-induced GHG emissions is a top priority for many communities around the world. From a different standpoint, communities are consistently under pressure to increase operational cost-efficiency savings, including the consideration of spatially consolidating public resource provision. This is the case for public schools, where governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. This study quantifies the travel burdens of school consolidation/closure policy through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. Specifically, we quantify how these closures changed the accessibility landscape for the elementary school aged population by using spatial availability, a novel singly-constrained accessibility measure. Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property. The analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorzied and non-motorized modes. System-wide impacts resulting from the closures, such as potential GHG emission increases and walk minutes decreases, are detailed. Overall, spatial availability decreased in the city. Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pay the price. This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications associated with policy changes.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---

@@

<!--- 2016 and 2011 sp_avail is calculated in 'test-doc-2016run', LOAD THE BELOW! -->
```{r, include=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggspatial) # Spatial Data Framework for ggplot2
library(here) # A Simpler Way to Find Your Files
library(patchwork) # The Composer of Plots
library(scales) # Scale Functions for Visualization
library(sf) # Simple Features for R
library(stplanr) # Sustainable Transport Planning
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap) # Thematic Maps
library(units) # Measurement Units for R Vectors
```

@@

<!--
Additional methods considerations: circuity of trip by mode? Schoolbus is not as direct of a trip as car/taxi/bikeORwalk. Maybe we consider 3 modes instead of 2? The r5r travel time for car/taxi as well as walk/cycle can be used as is, but the travel time by school bus can be increased based on a trip distance factor (from TTS)?

RESULTS an alternative layout idea?:
1. Multimodal spatial availability aggregated at the DA for each year and for each mode. Each DA is ~200 parcel points with their low-income and other-income spatial availability per mode; both these are summed together along with the rest of the parcel points per da. (6 plots - 3 modes per 2011 and 3 modes per 2016). 
2. Income-specific spatial availability aggregated at the DA for each population type. Similarly, each DA is ~200 parcel points with 3 mode-specific spatial availability per income-group; all three are summed together along with the rest of the parcel points per da.  (4 plots - 2 population-groups per 2011 and 2 population-groups per 2016).
3. EXAMINING THE INTERMEDIATES: the change in spatial availability per focused communities:
3.a. Hamilton-Central+Dundas experienced the majority of school closures, also have the highest LICO-AT as well as motorized transport. Glanbrook+StoneyCreek experienced the majority of new schools (two), also have low LICO-AT as well as non-motorized transport. So. 2 columns.  First column SA By Mode. SA By Mode in 2016 for HC+Dundas, then SA By Mode in 2016 for Glan+SC, then SA By Mode in 2016 Overall. Second column SA By Pop. SAY by LICO or not LICO in 2016 for HC+Dundas, then for GLAN+SC, then for overall. Showing SA for 2016 with brackets that say (+ or - X% to indicate difference from 2011).  
3.b. Examining the change in travel time per focused communities: same bar graph style as in 3.a. Will show have non-motorized travel time decreased in Ham-central, esp for LICO(?)
3.c. Change in emissions: same bar graph style as in 3.a. but for emissions. Will show have emissions increased, esp in Ham-central and LICO(?)
4. USING SPATIAL AVAILABILITY FOR PLANNING. Equity? We want to plan for school-seat availability _for_ active modes. Especially in areas with currently high LICO-AT. So lets plot spatial availability per non-motorized-LICO-AT capita. We'd like to see 1.2 SA per capita everywhere. We need improvements to the population, opportunities, and non motorized transport network (relative to all other modes/population groups) is the key point. -->

@@

In this study, we focus on the spatial travel implications of school closure and consolidation policies. The quantification of lost potential active transportation trips and associated carbon emissions due to school closures and the unequal impacts attributed to the increased reliance on a motorized commute is scarce in the literature. In this way, our contribution is both methodological and empirical. We quantify the accessibility changes before- and after the first wave of school closure/consolidations implemented in Hamilton, a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in Ontario, Canada [@governmentofcanadaProfileTableCensus2022].

In 2013, the City of Hamilton's largest English school board the Hamilton-Wentworth District School Board (HWDSB), which is the largest English school board in the city, released their Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013]. To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subject to accommodation strategies, including a "Accommodation Review" whose outcome is possibly closure/consolidation [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014]. These decisions were in part motivated by operational savings based on projected reductions in student population and under-utilized school capacity [@craggsAxeCouldFall2012]. However, the case of Hamilton and the HWDSB is unlike others in the province: the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014]. Though the Accommodation Review guidelines outline a community-focused process to deliberate a school's future viability to students, the community and the economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], concerned residences felt a fulsome breakdown of costs to repair schools were not made consistently available [@kleinhuisSeriousConcernsSchool2013]. Further, others pointed to flawed incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013]. Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018]. In a backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards (as they no longer have the ability to levee property taxes) [@mackenzieBlueprintFixOntario2018].

@@

 >CUT? The remainder of the paper is organized as follows. Section 2 details the data and methods and Section 3 showcases the results of the impact that the school consolidation policy had on Hamilton's school spatial availability landscape. Section 4 discusses the results and estimates potential walk minutes lost, where trip-related GHG emissions may have increased, and equity implications for the policy. Section 5 concludes the paper with key remarks.

@@

Equation (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) describes the estimated OTGC for public schools and catholic schools respectively. School status and OTGC summed by catchment is also visualised in @fig-Fig1. See the Table \ref{Tab:TabA1} in the Appendix for the associated regression results.

@@

knitr::include_graphics(paste0(here::here(),"/Figures/catchment_schools_plot.png"))
```

## Students, residential locations and low-income prevelance
Second, student population and proportion of household low-income after tax (LIM-AT) prevalence is retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]. The census releases population data by age group category in 5-year increments. The LIM-AT prevelance for households with children under 18 (there is no LIM-AT prevelance tabulated for households with exclusively elementary aged children) is retrained. LIM-AT is refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017]. These census variables are taken at the Dissemination Area (DA) level, which is the finest geographic unit for which census data is made publicly available. Census data is retrieved using the {cancensus} R package [@von_Bergmann_Shkolnik_Jacobs_2021]. DAs are designed by Statistics Canada with the aim of population uniformity hence DAs greatly vary in area but represent between approximately 400 and 700 (1st and 3rd quartile) in total population. The population aged 5 to 14 and the proportion of LIM-AT are visualised in the first and last row in @fig-Fig2.

The third data source consists of centroids of residential parcels in 2011 and 2016 that are used to represent residential locations [@Teranet_2009]. There are 134,340 and 139,467 residential locations in 2011 and 2016 respectively, representing `r cen_data_2011_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` and `r cen_data_2016_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` households (according to the Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]). As the number of children per residential location is unavailable publicly, the average number of 5 to 14 year olds for each DA is divided by the number of residential households in that DA. Then the potential number of children at each residential location is calculated by multiplying this DA rate. Due to the proprietary nature of the parcel data, the middle row in @fig-Fig2 visualises an aggregation of the information: the average rate of 5-14 year olds per residential household at the DA level.

@@

It is assumed that children can go to any elementary school, but certain schools are more likely to be attended depending on their proximity to their residential location. Based on this assumption, the fourth data source is empirical data regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018]. The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario every 5 years. With a target 5% sampling rate, the survey is expanded to be representative at the traffic analysis zone (TAZ) level of geography. TAZ are spatial units created for the purpose of the TTS; a few DAs typically nest within each TAZ. The trip-level travel data extract for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private care passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home to school in 2011 and 2016 respectively. {r5r} is used to estimated motorized and non-motorized travel from all TAZ centroids to all TAZ centroids and these estimated travel times are tied to the OD trips. The travel times and the intensity of modal home-to-school flows are visualised in @fig-Fig3 along with the boundaries of the TAZ and DAs. As can be seen in @fig-Fig3, not all TAZs capture an elementary school trip by both modes. For this reason, the proportion of modal-split is aggregated at the community level. Hamilton (Ancaster `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).

@@

The fifth data source consists of estimated travel time matrices from all residential parcel locations to all elementary schools for both motorized (car) and non-motorized (walking) travel. The travel times are estimated using the {r5r} [@Pereira2021r5r] package, which interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022]. For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrik [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the Hamilton boundary). Further, walk travel times that are only less than 27 min minutes were retained. This travel time is approximately equivalent to 1.6 km which is the minimum distance that is required from a school within-catchment to qualify for a school bus [@HWDSB_2019]. <!--put in statistics here-->

These travel time matrices along with the origin-destination flows are used to create modal trip length distribution (TLD) functions. A TLD is a useful technique to calibrate impedance functions as they are a representation of the likelihood that a proportion of trips are taken at a specific travel cost [@horbachov_theoretical_2018; @batista_estimation_2019]. The empirical TLD is fit to a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015]. Based on goodness-of-fit criteria and diagnostics (included in the Appendix), the gamma and exponential distributions are selected for the motorized and non-motorized modal distributions respectively. The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016). The exponential distribution is defined by the rate ($\beta$) parameter of `r round(exp2011_nonmotor$estimate[1], 3)` (2011) and `r round(exp2016_nonmotor$estimate[1], 3)` (2016). The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4. <!--put in equations for the functions-->

```{r}
#| label: fig-Fig4
#| fig-cap: "Empirical and theoretical motorized and non-motorized impedance functions."
#| echo: false

knitr::include_graphics(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"))
```

@@

knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"))
```

The motorized and non-motorized spatial availability $V_i$ for both 2011 and 2016 is represented in the top half of @fig-Fig6. Though at different magnitudes, both populations using non- and motorized modes have higher spatial availability when more proximate to schools. Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities. Trends appear similar for both 2011 and 2016: proximate populations using motorized modes benefit from a wider-range in high spatial availability than non-motorized modes. When $V_i^m$ is divided by the number of mode-using students in that DA, $v_i^m$ is derived (bottom half of @fig-Fig6). A similar spatial distribution seen in $V_i^m$ can be seen in $v_i^m$ plots, but as a rate of student population. If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (i.e., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes, and notably only for non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and certain pockets of Hamilton Central. The majority of schools closed were within Hamilton Central, so non-motorized populations in DAs within Hamilton Central were the most negatively impacted by losses in availability while certain motorized populations captured these losses as availability gains.

@@

knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"))
```

@@

<!--Maybe worth adding a plot of GHG emissions or change in tt?-->

>>>>

---
title: "The cost of school consolidations: impacts of reduced accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
linenumbers: true
numbersections: true
format:
  nature-pdf: default
editor: source
execute: 
  freeze: auto
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: McMaster University SEES
    footnote: "Corresponding Author"
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliation: University of Toronto
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca "
    affiliation: McMaster University CE
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: McMaster University SEES
address:
  - code: McMaster University SEES
    address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - code: McMaster University CE
    address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - code: University of Toronto
    address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"  
abstract: | 
  | The reduction of mobility-induced GHG emissions is a top priority for many communities around the world. From a different standpoint, communities are consistently under pressure to increase operational cost-efficiency savings, including the consideration of spatially consolidating public resource provision. This is the case for public schools, where governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. This study quantifies the travel burdens of school consolidation/closure policy through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. Specifically, we quantify how these closures changed the accessibility landscape for the elementary school aged population by using spatial availability, a novel singly-constrained accessibility measure. Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property. The analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorzied and non-motorized modes. System-wide impacts resulting from the closures, such as potential GHG emission increases and walk minutes decreases, are detailed. Overall, spatial availability decreased in the city. Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pays the price. This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications associated with policy changes.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---

@@

```{r, include=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggspatial) # Spatial Data Framework for ggplot2
library(here) # A Simpler Way to Find Your Files
library(patchwork) # The Composer of Plots
library(scales) # Scale Functions for Visualization
library(sf) # Simple Features for R
library(stplanr) # Sustainable Transport Planning
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap) # Thematic Maps
library(units) # Measurement Units for R Vectors
```

@@

@@

In this study, we focus on the spatial travel implications of school closure and consolidation policies. The quantification of lost potential active transportation trips and associated carbon emissions due to school closures and the unequal impacts attributed to the increased reliance on a motorized commute is scarce in the literature. In this way, our contribution is both methodological and empirical. We use a novel singly-constrained accessibility measure to quantify changes before- and after the first wave of school closure/consolidations implemented in Hamilton, a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in Ontario, Canada [@governmentofcanadaProfileTableCensus2022].

In 2013, the City of Hamilton's largest English school board the Hamilton-Wentworth District School Board (HWDSB), which is the largest English school board in the city, released their Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013]. To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subject to accommodation strategies, including a "Accommodation Review" whose outcome is possibly closure/consolidation [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014]. These decisions were in part motivated by operational savings based on projected reductions in student population and under-utilized school capacity [@craggsAxeCouldFall2012]. However, the case of Hamilton and the HWDSB is unlike others in the province: the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014]. Though the Accommodation Review guidelines outline a community-focused process to deliberate a school's future viability to students, the community and the economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], concerned residences felt a fulsome breakdown of costs to repair schools were not made consistently available [@kleinhuisSeriousConcernsSchool2013]. Further, others pointed to flawed incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013]. Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018]. With the backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards as they no longer have the ability to levee property taxes [@mackenzieBlueprintFixOntario2018].

@@

@@

Equation (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) describes the estimated OTGC for public schools and catholic schools respectively. School status and OTGC summed by catchment is also visualised in @fig-Fig1. See the Table <!--\ref{TabA1-OTGC}--> in the Appendix for the associated regression results. <!-- shall we explain further how we valuated this model? I don't think its entirely necessary. But the code is here: https://github.com/soukhova/hamONdests/blob/master/data-raw/01-school-data-preparation.Rmd (Antonio is a collaborator) -->

@@

knitr::include_graphics(paste0(here::here(),"/Figures/catchment_schools_plot.png"))
```
<!--perhaps in the legend of this map, I add the number of schools in brackets (for "school status between 2011 and 2016")? And the number of schools on the map in each box? what do you think? I'll also make the font on the panels bigger-->

## Students, residential locations and low-income prevelance
Second, student population and proportion of household low-income after tax (LIM-AT) prevalence is retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]. The census releases population data by age group category in 5-year increments. The LIM-AT prevalence for households with children under 18 (there is no LIM-AT prevalence tabulated for households with exclusively elementary aged children) is retrained. LIM-AT is refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017]. These census variables are taken at the Dissemination Area (DA) level, which is the finest geographic unit for which census data is made publicly available. Census data is retrieved using the {cancensus} R package [@von_Bergmann_Shkolnik_Jacobs_2021]. DAs are designed by Statistics Canada with the aim of population uniformity hence DAs greatly vary in area but represent between approximately 400 and 700 (1st and 3rd quartile) in total population. The population aged 5 to 14 and the proportion of LIM-AT are visualised in the first and last row in @fig-Fig2.

The third data source consists of centroids of residential parcels in 2011 and 2016 that are used to represent residential locations [@Teranet_2009]. There are 134,340 and 139,467 residential locations in 2011 and 2016 respectively, representing `r cen_data_2011_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` and `r cen_data_2016_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` households (according to the Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]). As the number of children per residential location are publicly unavailable, the average number of 5 to 14 year olds for each DA is divided by the number of residential households in that DA. Then the potential number of children at each residential location is calculated by multiplying this DA rate. <!-- NOTE: We consider the parcel centroids instead of DA centroids to more precisely calc. travel times. Since we do not know parcel student population, we assume a uniform spatial distribution of population across each DA. This assumption is a limitation, but it is better than what's typically done in studies that only consider the centroid of the census track. --> Due to the proprietary nature of the parcel data, the middle row in @fig-Fig2 visualises an aggregation of the information: the average rate of 5-14 year olds per residential household at the DA level.

@@

It is assumed that children can go to any elementary school, but certain schools are more likely to be attended depending on their proximity to their residential location. Based on this assumption, the fourth data source is empirical data regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018]. The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario every 5 years. With a target 5% sampling rate, the survey is expanded to be representative at the traffic analysis zone (TAZ) level of geography. TAZ are spatial units created for the purpose of the TTS; a few DAs typically nest within each TAZ. The trip-level travel data extract for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private care passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home to school in 2011 and 2016 respectively. The travel times for motorized and non-motorized travel are estimated using the {r5r} [@Pereira2021r5r] package, which interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022]. Motorized and non-motorized travel times are estimated for all TAZ centroids to all TAZ centroids and these estimated travel times are tied to the associated OD trips. The travel times and the intensity of modal home-to-school flows are visualised in @fig-Fig3 along with the boundaries of the TAZ and DAs. As can be seen in @fig-Fig3, not all TAZs capture an elementary school trip by both modes. For this reason, the proportion of modal-split is aggregated at the community level. Hamilton (Ancaster `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`). <!--Do we maybe need to discuss how this lack of data for each TAZ impacted the calculations?-->

@@

The fifth data source consists of estimated travel time matrices from all residential parcel locations to all elementary schools for both motorized (car) and non-motorized (walking) travel. The travel times are estimated using the {r5r} package [@Pereira2021r5r]. For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrik [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the Hamilton boundary). Further, walk travel times that are only less than 27 min minutes were retained. This travel time is approximately equivalent to 1.6 km which is the minimum distance that is required from a school within-catchment to qualify for a school bus [@HWDSB_2019]. Estimated trip-weighted motorized travel times are on average 16.5 minutes in 2011 and 16.8 minutes in 2016 (summary statistics for both years are approximately min: 0.5, Q1: 11.0, Q2: 16.0, Q3: 21.0, max: 60.0). Estimated trip-weighted non-motorized travel times are on average 18.0 minutes in 2011 and 18.1 minutes in 2016 (summary statistics for both years are approximately min: 0.5, Q1: 13.0, Q2: 19.0, Q3: 24.0, max: 27.0).
<!--to calc. the summary statistics i loaded "OD_2011" and "OD_2016" from data-products (big files) and ran: quantile(OD_2011|>filter(mode=="w" & travel_time_p50 <= 27)|>pull(travel_time_p50), weights=OD_2011|>filter(mode=="w")|>pull(kid_popadj)) 
and 
mean(OD_2011|>filter(mode=="w" & travel_time_p50 <= 27)|>pull(travel_time_p50), weights=OD_2011|>filter(mode=="w")|>pull(kid_popadj))
-->

These travel time matrices along with the origin-destination flows are used to create modal trip length distribution (TLD) functions. A TLD is a useful technique to calibrate impedance functions as they are a representation of the likelihood that a proportion of trips are taken at a specific travel cost [@horbachov_theoretical_2018; @batista_estimation_2019]. The empirical TLD is fit to a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015]. Based on goodness-of-fit criteria and diagnostics <!-- should I include the diagonostic plots in the appendix? q-q, etc..? I don't think its entirely necessary as the empirical is graphed over the theoretical line in the following figure-->, the gamma and exponential distributions are selected for the motorized and non-motorized modal distributions respectively. The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016). The exponential distribution is defined by the rate ($\beta$) parameter of `r round(exp2011_nonmotor$estimate[1], 3)` (2011) and `r round(exp2016_nonmotor$estimate[1], 3)` (2016). The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4. For reference, the gamma distribution and the exponential distribution function are displayed in the following Equation (\ref{eq:exp-dist})) and (\ref{eq:gamma-dist})) where $x$ is $c_{ij}$:

```{=tex}
\begin{equation}
\label{eq:exp-dist}
f(x) = \beta e ^{-\beta x}
\end{equation}
```

```{=tex}
\begin{equation}
\label{eq:gamma-dist}
\begin{aligned}
f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty \\
\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx
\end{aligned}
\end{equation}
```

```{r}
#| label: fig-Fig4
#| fig-cap: "Empirical (grey bars) and theoretical (blue curves) motorized and non-motorized impedance functions. Motorized theoretical impedance function is based on the gamma distribution function and non-motorized on the expoential distribution function"
#| echo: false

knitr::include_graphics(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"))
```
<!--maybe bigger font here?-->

@@

knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"))
```
<!--I'll make this plot longer-->
The motorized and non-motorized spatial availability $V_i$ for both 2011 and 2016 is represented in the top half of @fig-Fig6. Though at different magnitudes, both populations using non- and motorized modes have higher spatial availability when more proximate to schools. Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities. Trends appear similar for both 2011 and 2016: proximate populations using motorized modes benefit from a wider-range in high spatial availability than non-motorized modes. When $V_i^m$ is divided by the number of mode-using students in that DA, $v_i^m$ is derived (bottom half of @fig-Fig6). A similar spatial distribution seen in $V_i^m$ can be seen in $v_i^m$ plots, but as a rate of student population. If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (i.e., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes, and notably only for non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and certain pockets of Hamilton Central. The majority of schools closed were within Hamilton Central, so non-motorized populations in DAs within Hamilton Central were the most negatively impacted by losses in availability while certain motorized populations captured these losses as availability gains.

@@

knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"))
```
<!--The colours combos I choose here are absolutely ugly ;( but besides that, do you agree that V_im and v_im should use different colour palettes?-->

@@

<!--Antonio, what do you think, maybe worth adding a plot of GHG emissions or change in tt? This will provide us some higher level summary stats. Moataz asked about this as well regarding the figures in the Results section-->

<!--Antonio, what do you think, add a table with summaries of LIMAT, Kid pop, V_im change, v_im change, total travel time change, and emission changed by community and mode? We should discuss to decide what's the best presentation.-->
```{r}
# SAvail_im_COMM_changes |> filter(mode == "motorized") |> st_drop_geometry() |> 
#   dplyr::select(c("COMMUNITY",
#                   "prev_LIMAT_2016", 
#                   "c_kid_perc", 
#                   "c_V_im_perc", 
#                   "c_v_im_perc", 
#                   "emissions_2011",
#                   "emissions_2016"
#                   #"c_emissions_perc",
#                   # "emissions_perkid_2011",
#                   # "emissions_perkid_2016",
#                   #"c_emissions_perkid_perc" 
#                   )) |> arrange(desc(prev_LIMAT_2016)) 
```
